---
title: "Lab2 - The `grid` Graphics System"
author: Mathias L. Jeppesen
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
  prettydoc::html_pretty:
    theme: tactile
---

# Setup

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  dev.args = list(png = list(type = "cairo"))
)
```

Loading packages.
```{r message = FALSE}
library(ggplot2)
library(grid)
library(tidyverse)
```

# The data and questions of interest

Data shows rows of incidents handled by the Police.

We wish to answer questions about youth offending (up to age 19) regarding incidence rate overall as well as ratio between male and female offenders over time. We also want to investigate whether our conclusions change if we stratify by crime type.

```{r}
crime <- read.csv("nzpolice-proceedings.csv")
head(crime)
youth <- subset(crime, Age.Lower == 15)
```

# Data visualisations and questions

## Question 1

```{r}
youthSex <- table(youth$SEX)/nrow(youth)
female <- "#E46C0A"
male <- "#0070C0"
```

We solve this task by creating a function for generating a `grobTree` using information from data. The function is made to help produce plots in questions 3, 5 and 8 as well (we need a grob(Tree) for the implementation in the challenge question 8).

```{r}
youthSexData <- as.data.frame(youthSex) %>% 
  setNames(c("sex", "rate"))

barRatio_grobTree <- function(
    data = youthSexData,
    bar_height = unit(1, "cm"),
    cols = list(female = female, male = male)) {
  data <- data %>% 
    dplyr::mutate(
      label = c(paste0(sex, " ", round(rate * 100), "%"))
    )
  femaleData <- data %>% 
    dplyr::filter(sex == "Female")
  maleData <- data %>% 
    dplyr::filter(sex == "Male")
  
  fbar <- rectGrob(x = 0,
                   hjust = 0,
                   width = unit(femaleData$rate, "npc"),
                   height = bar_height,
                   gp = gpar(col = cols$female, fill = cols$female))
  mbar <- rectGrob(x = 1,
                   hjust = 1,
                   width = unit(maleData$rate, "npc"),
                   height = bar_height,
                   gp = gpar(col = cols$male, fill = cols$male))
  
  ftext <- textGrob(x = unit(2, "mm"),
                    label = femaleData$label,
                    hjust = 0,
                    gp = gpar(fontface = "bold"))
  mtext <- textGrob(x = unit(1, "npc") - unit(2, "mm"),
                    label = maleData$label,
                    hjust = 1,
                    gp = gpar(fontface = "bold"))
  
  grobTree(fbar, mbar, ftext, mtext)
}
```

Produce the plot using the function.

```{r, fig.height = 1}
grid.draw(barRatio_grobTree())
```

## Question 2

To solve this task, we just use the `barRatio_grobTree` function inside a viewport of appropriate size.

```{r, fig.height = 1}
pushViewport(viewport(width = unit(1, "npc") - unit(2, "cm")))
grid.draw(barRatio_grobTree())
upViewport()
```

## Question 3

We define a data set similar to in question 1 but in long format with a pair of rows for females and males for each year. We then create the plot using the `barRatio_grobTree` function defined in question 1 within appropriate viewports.

```{r}
youthSexYear <- t(apply(table(substr(youth$Date, 1, 4), youth$SEX), 1, 
                        function(x) x/sum(x)))

## Creating data in long format
youthSexYearData <- as.data.frame(t(youthSexYear)) %>%
  dplyr::mutate(sex = row.names(.)) %>% 
  tidyr::pivot_longer(cols = num_range(prefix = "", range = 2014:2022),
                      names_to = "year",
                      values_to = "rate") %>% 
  dplyr::arrange(desc(year)) %>%
  dplyr::mutate(id_year = dplyr::cur_group_id(), .by = year)

plot9bars <- function(data = youthSexYearData) {
  id_years <- unique(data$id_year)
  num_years <- length(id_years)
  
  grid.newpage()
  pushViewport(
    viewport(
      layout = grid.layout(nrow = num_years),
      width = unit(1, "npc") - unit(1, "in") - unit(1, "cm"),
      x = unit(1, "in"),
      height = unit(1, "npc") - unit(2, "cm"),
      y = unit(1, "cm"),
      just = c("left", "bottom")
    )
  )
  
  for (i in id_years) {
    data_1year <- data %>%
      dplyr::filter(id_year == i)
    bar <- barRatio_grobTree(
      data = data_1year,
      bar_height = convertHeight(unit(.08, "npc"), "inches")
    )
    pushViewport(
      viewport(
        layout.pos.col=1, layout.pos.row=i
      )
    )
    grid.draw(bar)
    upViewport()
  }
  popViewport(0)
  
  pushViewport(
    viewport(
      layout = grid.layout(nrow = num_years),
      width = unit(1, "in"),
      x = unit(1, "in") - unit(2, "mm"),
      height = unit(1, "npc") - unit(2, "cm"),
      y = unit(1, "cm"),
      just = c("right", "bottom")
    )
  )
  for (i in id_years) {
    year <- data %>%
      dplyr::filter(id_year == i) %>% 
      dplyr::pull(year) %>% 
      unique()
    pushViewport(viewport(layout.pos.col=1, layout.pos.row=i))
    grid.text(year, x = 1, hjust = 1)
    upViewport()
  }
  popViewport(0)
}
```

Generate plots with default and specified `fig.height/width` respectively, using the new `plot9bars` function.

```{r}
plot9bars()
```

```{r, fig.height=3, fig.width=5}
plot9bars()
```

This plot gives us information about our first question of interest about the ratio of female and male offenders across time. The bars show that the proportion of female offenders has been monotonically decreasing in the years 2018-2022, where the proportion of females went from 29% of offenders to 24%.

## Question 4

Define the data given in the assignment.

```{r}
youth$Abbrev <- 
  gsub(",", "",
       unlist(lapply(strsplit(as.character(youth$ANZSOC.Division), " "),
                     function(x) x[1])))
youthTrendType <- as.data.frame(table(youth$Date, youth$Abbrev))
youthTrendType$Date <- as.Date(youthTrendType$Var1)
## Order crime types
types <- table(youth$Abbrev)
newlevels <- names(types)[order(types, decreasing=TRUE)]
youthTrendType$Type <- factor(youthTrendType$Var2, levels=newlevels)

abb <- unique(youth[c("Abbrev", "ANZSOC.Division")])
```

We create the facetted line plot using `ggplot2`.

```{r}
facet_plot <- function(data = youthTrendType, scales = "fixed", ...) {
  data %>% 
    ggplot(aes(x = Date, y = Freq, ...)) +
    geom_line(show.legend = FALSE) +
    facet_wrap(~ Type, scales = scales)
}
facet_plot()
```

This plot tells nothing about gender differences in crime offence. However, the plot shows time trends for different crime types. It seems that for almost all types of crime, there is a downward trend in the frequency of offences. The only type of crime that is clearly not on a downward trend is the "Dangerous" (`r abb[abb$Abbrev == "Dangerous", ]$ANZSOC.Division`) crimes.

However, this plot does have an issue since the `facet_wrap` uses `scales = "fixed"`, meaning that the y-axis scale is the same for all crime types. This makes it hard to see what is happening with the crime types that are not as frequent as fx. theft. To investigate further, we could explore the use of `scales = "free"`.

## Question 5

To solve this exercise, we can simply create viewports to plot inside. We can assign the `ggplot` object as a value in our environment and then plot it inside the viewport with `newpage = FALSE`. Then, we can plot the bar with ratios on top. A viewport is not strictly necessary here, but makes the function I created earlier work with default behavior.

```{r}
fac <- facet_plot()
pushViewport(
  viewport(
    height = unit(1, "npc") - unit(1, "cm"),
    y = 0,
    just = c("centre", "bottom")
  )
)
plot(fac, newpage = FALSE)
upViewport()

pushViewport(
  viewport(
    height = unit(1, "cm"),
    y = unit(1, "npc"),
    just = c("centre", "top")
  )
)
grid.draw(barRatio_grobTree())
upViewport()
```

## Question 6

We use the `grid` package to define a *grob* (graphical object) using the `xGrob` functions. In this case, we create a transparent blue box with `rectGrob` and then add it to each facet of the `ggplot` using the `grid_panel` function from `gggrid`, which adds the grob to each facet.

```{r}
transBlueBox <- rectGrob(
  width = unit(.15, "npc"),
  x = 1,
  hjust = 1,
  gp = gpar(col = male, fill = male, alpha = 0.35))

facet_plot() +
  gggrid::grid_panel(transBlueBox)
```

## Question 7

To solve this task, we can make use of the fact that `gggrid::grid_panel` can take a function producing a grob rather than a static grob as input. In this case, since we want to color the part of the graph from 2022, we create a data set consisting of only values from this period. We then create a function which creates a `rectGrob` as a function of input data, using ranges to define the size of the rectangle. We then input this function `add_bar` in `gggrid::grid_panel`, which uses the function within each panel. Thus, within each panel, data will consist of observation only from year 2022 and that type of crime. We can then simply find the ranges of the x- and y- values to define the area of the rectangle.

```{r}
youthTrendType2022 <- youthTrendType %>% 
  dplyr::filter(lubridate::year(Date) >= 2022)

add_bar <- function(data, coords) {
  max_y <- max(coords$y)
  min_y <- min(coords$y)
  
  max_x <- max(coords$x)
  min_x <- min(coords$x)
  diff_x <- max_x - min_x
  
  rectGrob(
    x = min_x,
    width = diff_x,
    hjust = 0,
    y = min(coords$y),
    height = max_y - min_y,
    vjust = 0,
    gp = gpar(col = male, fill = male, alpha = .35)
  )
}

facet_plot() +
  gggrid::grid_panel(add_bar,
                     data = youthTrendType2022)
```

This plot, like in question 4 and 6 does not help us answer questions regarding gender differences in offending. In question 4 we discussed how the plot can be used to identify a downward trend over time for most crime types except for "Dangerous" (`r abb[abb$Abbrev == "Dangerous", ]$ANZSOC.Division`) crimes.

This plot however highlights data from year 2022, which makes us notice an upward trend in data from this year alone for all crime types. This is something that was not easily "catchable" in the plot from question 4. So this plot does help us understand what is happening from year 2022 and onward.

The plot still suffers from the fact that not-so-frequent crime types are hard to see, so below I have made the same plot but with `scales = "free_y"` to better be able to investigate the low-frequency crime types.

```{r}
facet_plot(scales = "free_y") +
  gggrid::grid_panel(add_bar,
                     data = youthTrendType2022)
```

This plot shows that even for the low-frequency crime types there is the same downward trend in most of data but with a small increase from year 2022 and onwards.

## Challenge

Define data as in assignment to use for plot:

```{r}
youthTrendTypeSex <- as.data.frame(table(youth$Date, youth$Abbrev,
                                         youth$SEX))
youthTrendTypeSex$Date <- as.Date(youthTrendTypeSex$Var1)
youthTrendTypeSex$Type <- factor(youthTrendTypeSex$Var2, levels=newlevels)
youthTrendTypeSex$Sex <- factor(youthTrendTypeSex$Var3)
```

Like in question 7, we create a function with arguments `data` and `coords` that we input to `gggrid::grid_panel`. In this case, we create a summary of the proportion of female and male offenders within each panel (crime type) and use these summaries to create bars on the right to visualise this. We also add white horizontal lines at break points 0.1 to 0.5 to make it more easy to see the distribution of men vs. women in each panel.

To create such a collection of grobs (2 rectangles to form the bar with proportions and a segment with white lines), we use the `gTree` function and specify all 3 grobs as `children` using a `gList`.

To also add a bar with the overall proportions to the right of the plot, we simply utilise viewports as in question 5, but now using the `angle` argument to create a vertical bar. We also extract the colors used as a default from `ggplot` to color our groups and use them for our bar.

```{r}
add_rate <- function(data, coords) {
  rateData <- data %>%
    summarise(offences = sum(y), .by = colour) %>%
    mutate(rate = offences / sum(offences))
  
  bar1 <- rectGrob(
    x = 1,
    width = unit(1, "mm"),
    hjust = 1,
    y = 0,
    height = rateData[1, ]$rate,
    vjust = 0,
    gp = gpar(col = rateData[1, ]$colour,
              fill = rateData[1, ]$colour)
  )
  bar2 <- rectGrob(
    x = 1,
    width = unit(1, "mm"),
    hjust = 1,
    y = 1,
    height = rateData[2, ]$rate,
    vjust = 1,
    gp = gpar(col = rateData[2, ]$colour,
              fill = rateData[2, ]$colour)
  )
  lines <- segmentsGrob(x0 = unit(1, "npc"), x1 = unit(1, "npc") - unit(1, "mm"),
                        y0 = seq(0.1, 0.5, 0.1), y1 = seq(0.1, 0.5, 0.1),
                        gp = gpar(col = "white", fill = "white"))
  gTree(children = gList(bar1, bar2, lines))
}

sexFacetPlot <- function(data = youthTrendTypeSex, scales = "fixed") {
  grid.newpage()
  p <- facet_plot(data = data,
                  color = Sex,
                  scales = scales) +
    gggrid::grid_panel(add_rate)
  
  pushViewport(
    viewport(
      x = 0,
      just = c(0, .5),
      width = unit(.9, "npc")
    )
  )
  plot(p, newpage = FALSE)
  upViewport()
  
  cols_in_plot <- scales::pal_hue()(2)
  cols_as_list <- as.list(cols_in_plot) %>% setNames(c("female", "male"))
  
  pushViewport(
    viewport(
      x = unit(1, "npc") - unit(.1, "npc") / 2,
      height = unit(.1, "npc"),
      width = convertUnit(
        unit(0.7, "npc"),
        unitTo = "in",
        axisFrom = "y",
        axisTo = "x"),
      angle = 90
    )
  )
  grid.draw(barRatio_grobTree(cols = cols_as_list,
                             bar_height = unit(6.5, "mm")))
  upViewport()
}
```

Use the new function to create the plot:

```{r}
sexFacetPlot()
```

This plot now helps us answer all questions of interest. Again, using free `scales` in the facets, I think we gain more insight in what is happening:

```{r, fig.width=12, fig.height=8}
sexFacetPlot(scales = "free_y")
```

# Overall summary

## The `grid package`

The `grid` package allows total customisation in creating *grobs* (graphical objects). They can either be drawn directly using `grid::grid.xx` functions or created as grobs with `grid::xxGrob` functions to be drawn later (with `grid::grid.draw`). This enables us to gather different grobs in a *grob tree* with `grid::gTree` and specifying grobs with a `grid::gList` as the `children` argument (alternatively use `grid::grobTree` when all components are grobs). See more in *question 1*.

### Using viewports

We can use `grid::viewport` with `grid::pushViewport` to draw whatever we want inside a certain area of our "full plotting area" and then go to a parent viewport with `grid::upViewport` (`grid::popViewport` navigates to parent viewport AND deletes the current viewport from the viewport tree). See *questions 2, 3, 5 and 8*.

Additional takeaways about viewports:

- **Note** this also works with `ggplot2` by assigning the `ggplot` object a value (fx. `p`) and then plotting it "inside the viewport" with `plot(p, newpage = FALSE)`. See *questions 5 and 8*.
- We can use the `grid::convertUnit` function to change the axis we are referencing (useful fx. when creating a viewport with `angle = 90/270`). See *question 8*.
- A strong way of using viewports is using `xscale` and/or `yscale` parameters in the `grid::viewport` function and the using the `grid::unit` function with `units = "native"` specified to make your grobs relate to the scale of the viewport.
- Use of `grid::grid.layout` as in *question 3* can help with many situations.

**Note** that `ggplot2` is built on `grid`. `ggplot2` uses viewports from `grid`. Fx. on a simple plot, it uses (nested) viewports for the background, axis text, axis ticks, etc. etc.

## Combining `grid` and `ggplot2` - The `gggrid` package

As mentioned above, viewports and ggplots integrate nicely using just `grid` and `ggplot2`. However, it's easy to combine elements of `grid` to `ggplot`s to create overlays on each facet/panel with `gggrid::grid_panel` or for each group using `gggrid::grid_group`.

These functions either take a static grob or a function that will create a grob (or a tree of grobs) within each panel/group. The function takes `data` and `coords` as an argument, and it will then use this function to draw using the function on only the data in that panel (stratified data). Note that the `coords` are transformed values of `x` and `y` (from the aesthetic mapping) to scale 0-1, so it can be plotted correctly in the scale of the viewport. `data` contains the untransformed data and includes columns of other aesthetics (fx. `colour`).

See more in *questions 6, 7 and 8*.

## Answering questions of interest

The proportion of female vs. male offenders is on a downward trend with the proportion of female offenders monotonically decreasing from 29% in 2018 to 24% in 2022, which is visible from the plot in question 3.

From question 4 (and 5) we see an overall downward in frequency of offences. We do however see an increase for a single crime type "Dangerous" (`r abb[abb$Abbrev == "Dangerous", ]$ANZSOC.Division`). Looking at the plot in question 6, it becomes apparent that across all crime types, there seems to be an increase from the start of 2022. This is further highlighted in question 7, where we use a more precise, data-driven method to show the increase from start of year 2022.

Looking at the plot from the challenge, we can comment on some of the same things as in the other plots, but those plots are more "specific" and perhaps a bit better at answering each question. This plot however gives an overview of both the proportion of female vs. male offenders through time as well as overall frequency across time and crime types. The plot also provides information not previously available by showing the proportions for each crime type.

